---
output:
  html_document:
    keep_md: yes
  md_document:
    variant: markdown_github
---

# initialize project 1

```{r echo = FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

```{r, echo = FALSE}
# my tutorial functions for text_icon and code_icon
source("helper_02_icons.R")
```

## start a private repo

If you haven't requested a GitHub student academic account, please do [https://education.github.com/pack](https://education.github.com/pack). On GitHub: 

- Create a repo called `me497-project-1-yourLastName`, replacing `yourLastName` with your actual last name 
- Make the repo private 
- Check the `Initialize this repository with a README` box 
- `Add gitignore` and choose `R` 
- A license is optional. 
- `Create repository`

Invite me to be your collaborator 

- In the repo, click on the `Settings` icon. 
- Click on the `Collaborators` tab 
- In the `Search by username` box, type my GitHub name `graphdr`, click `Add collaborator` to add me as a collaborator to your project 


In the repo main page, click `Clone or download` and copy that URL. We'll use it below.  


## link a new project to the repo 
  
In RStudio, create a new project: 

- RStudio *File > New Project > Version control > Git* 
- In the dialog  box, paste the repo URL, and edit the path so that the new project is a sub-directory of your course folder. 

Check your course folder. It should now contain a folder `me497-project-1-yourLastName`. Inside that folder you should find 

- `me497-project-1-yourLastName.Rproj`  
- `.gitignore`   
- `README.md`  

## setup the project directory tree

In your project directory, create new folders (sub-directories) for `data\`, `reports\`, `resources\`, `results\`, and `scripts\`. Earlier you created a `.Renviron` file. Copy and paste it to the project folder. Your directory should look like this: 

```
me497-project-1-YourLastName\
  |-- data\
  |-- reports\
  |-- resources\
  |-- results\
  |-- scripts\
  |-- me497-project-1-YourLastName.Rproj
  |-- .Renviron
  |-- .gitignore
  `-- README.md
```



 










---

[main page](../README.md)




