---
output: github_document
---

# Graphs 1 

Richard Layton    
2018-09-03 

"The goal of the first part of this book is to get you up to speed with the basic tools of data exploration as quickly as possible." 

- text pages 1--12 
- online [sections 3.1 and 3.2](http://r4ds.had.co.nz/data-visualisation.html#introduction-1) 

```{r setup}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, collapse=TRUE)
```

## 3.1.1

```{r}
library(tidyverse)
```

## 3.2.1

`mpg` is a data set in the tidyverse.

```{r}
mpg
```

## 3.2.2

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## 3.2.4  exercises

Do the exercises on your own for practice with the concepts. In this file, comment on what you see or what the exercise illustrates. 

1. Run ggplot(data = mpg). What do you see?

2. How many rows are in mpg? How many columns?

3. What does the drv variable describe? Read the help for ?mpg to find out.

4. Make a scatterplot of hwy vs cyl.

5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

## 3.3

With `class` mapped to the `color` aesthetic. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

With `class` mapped to the `size` aesthetic---produces a warning.  

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

With `class` mapped to the `alpha` aesthetic---produces a warning. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

With `class` mapped to the `shape` aesthetic---produces a warning. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Set the aesthetic properties of the geom manually, outside of `aes()`. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

## 3.3.1 exercises


Do the exercises on your own for practice with the concepts. In your version of this file, comment on what you see or what the exercise illustrates. 

1. What’s gone wrong with this code? Why are the points not blue?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

2. Which variables in mpg are categorical? Which variables are continuous? (Hint: type ?mpg to read the documentation for the dataset). How can you see this information when you run mpg?

3. Map a continuous variable to color, size, and shape. How do these aesthetics behave differently for categorical vs. continuous variables?

4. What happens if you map the same variable to multiple aesthetics?

5. What does the stroke aesthetic do? What shapes does it work with? (Hint: use ?geom_point)

6. What happens if you map an aesthetic to something other than a variable name, like aes(colour = displ < 5)?
