---
output: github_document
bibliography: "../resources/bib-etc/references-rr.bib"
csl: "../resources/styles/apa.csl"
link-citations: yes
--- 

# render all Rmd in reports

```{r}
library(knitr)
opts_knit$set(root.dir = ".")
opts_chunk$set(echo = TRUE)
```

## start 

We'll need the rmarkdpown package. 

```{r}
library(rmarkdown)
```

Name the path to the directory that contains the scripts we want to knit.

```{r}
path_to_Rmds <- "reports/"
```


## find the Rmd files 

Use `list.files()` to find all the Rmd scripts in that directory and save them in `report_filenames`, a vector of character strings.

```{r}
# find all files that end in .Rmd
report_filenames <- list.files(
	path = path_to_Rmds
	, pattern = ".Rmd$"
	, full.names = TRUE
)

report_filenames
```

## knit the Rmd scripts 

For each filename in `report_filenames`, apply the `render()` function one file at a time. The "one-file-at-a-time" process is managed by the `sapply()` function. 

```{r eval = FALSE}
# render each script
sapply(report_filenames, function(x) render(x))
```

Remove the html files

```{r}
unlink("reports/*.html")
```

