---
output:
  html_document:
    keep_md: yes
  md_document:
    variant: markdown_github
--- 

# tidy the data

```{r echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, results = 'hide', message = FALSE, warning = FALSE)
```

```{r echo = FALSE}
# functions for text_icon and code_icon
source("week_01/helper_02_icons.R")
```

How to use this tutorial 

- `r text_icon` *add text*: type the prose verbatim into the Rmd file 
- `r code_icon` *add code*: insert a code chunk, then transcribe the R code 
- `r knit_icon` *knit* after each addition. 
- *Install* packages one time only
- *Load* a package using `library()` every session

Packages used in this tutorial 

- readr
- tidyr 
- stringr
- dplyr 

## open a new Rmd script

Open a new Rmd file, and name it `02_calibr_data-tidy.Rmd`. Save it to the `scripts` directory.

Edit the YAML header:  

    ---
    title: "Load-cell calibration --- tidying the data"
    author: "your name"
    date: "date"
    output: html_document
    ---

Delete the rest of the pre-populated text. 

Insert knitr setup code. 

`r code_icon` 

    library(knitr)
    opts_knit$set(root.dir = '../')
    opts_chunk$set(echo = TRUE, collapse = TRUE)

Knowing the packages we'll be using, we can load them right away, near the top of the file.

`r code_icon`

```{r}
library(readr)
library(tidyr)
library(stringr)
suppressPackageStartupMessages(library(dplyr))
```

## plan the data reshaping

`r text_icon`

    # Plan the data reshaping

    For analysis, the data set should be in tidy (long) form, with every column a variable and every row an observation. The tidy form will have these variable names:

    - observ (observation number)
    - cycle (cycle number)
    - test\_pt (test point number and direction)
    - input\_lb (applied reference force)
    - output\_mV (sensor readings)

Learn R Markdown:

- `- ` (hyphen, space) creates an itemized list (bullet list). A line space above and below separating the list from other paragraphs is required. 
- To print an underscore in the Rmd prose we have to "escape" the character by writing `\_`. 

`r text_icon`

    In the reshaped data set, every mV reading will appear in its own row identified by cycle number (from the existing column name) and test point.  Thus the columns with *cycle* in their name  are the ones being gathered. 

The next code chunk finds the columns with `cycle` in their name. 

`r code_icon`

```{r}
# active data set
data_received <- read_csv('data/01_calibr_data_active-report.csv')
# extract the column names
all_col_names <- colnames(data_received)
```

Learn R 

- `colnames()` returns the column names (variable names) of the data frame 

Notes on reproducibility 

- Reading `01_calibr_data-active-report.csv` file is a consequence of DRY (don't repeat yourself). 
- For a full explanation, see the previous tutorial. 


`r code_icon`

```{r}
# TF, does the column name include "cycle"
detect_cycle_col <- str_detect(tolower(all_col_names), 'cycle')
```

Learn R 

- `tolower()` returns the vector all in lower case
- `str_detect()` (from `stringr`) returns a logical vector with TRUE for column each name containing `cycle`

`r code_icon`

```{r}
# extract indices of TRUE
is_a_cycle_col <- which(detect_cycle_col)
# examine
is_a_cycle_col
```

We know which columns contain the mV readings *for this data set*. However, those column numbers can change when we receive data updates. So we want the script to extract this type of information *from the data*. 

Writing readable code

- Don't hard code numbers --- use variables to reduce bugs, improve readability, and support reproducibility

Learn R 

- `which()` returns the indices of the position of TRUE
- Writing a variable on a line of its own, e.g., `is_a_cycle_col`, prints its value(s)
- `is_a_cycle_col` is a vector of integers. The number in brackets `[1]` is the index of the first element in the row of output. 

## check yourself

Confer with a neighbor

1. What are the indices to the cycle columns? 

## reshape the data to long form

We will gather the cycle numbers (the *names* of the original columns) in one new column and gather the mV readings (the *data values* in the original columns) in another new column. The `gather()` function from `tidyr` arranges each cycle and reading side by side in a new observation row. 

The new data frame has as many rows as there are observations in the original table. 

The columns "not gathered"" remain, e.g., `r all_col_names[-is_a_cycle_col]`, with their entries copied into the new rows, maintaining the relationships described in the original data set.

`r text_icon`

    # Reshape the data to long form

`r code_icon`

```{r}
# gather the columns of mV readings
long_data <- data_received %>%
	gather(cycle, output_mV, is_a_cycle_col) 
```

Learn R

- We can read this code as, start with `data_received`, *then* gather the variables (columns) designated by `is_a_cycle_col` into two new columns named `cycle` and `output_mV`, and assign the result to `long_data`. 
- `cycle` gathers the column names, e.g., `r all_col_names[is_a_cycle_col]`
- `output_mV` gathers the column data (mV), e.g., `r sprintf("%.1f", data_received[is_a_cycle_col][[2]])`, etc. 

`r code_icon`

```{r}
# examine its structure
glimpse(long_data) 

# examine the summary statistics of each column
summary(long_data) 
```

## check yourself

Confer with a neighbor.

1. The total number of observations is?
2. The total number of measured variables is? 

`r text_icon`

    I have the columns I expected.

    All the NA values are in the mV readings column and can be safely deleted because they represent readings never taken (per the ANSI/ISA standard). 

    Result is the first tidy form, where every row is an observation and every column a variable.  

`r code_icon`

```{r}
# omit NAs
tidy_data <- long_data %>%
	filter(!output_mV %in% NA)

# examine
glimpse(tidy_data)
```

Learn R

- `filter()` (from `dplyr`) returns all rows for which the logical expression is TRUE (deleting rows for which the expression is FALSE)
- `%in%` returns a logical vector indicating if the left operand (`!output_mV`) matches the right operand (`NA`)
- `!` is logical negation (NOT), so we can keep all the rows that are "not NA"

## add observation numbers

`r text_icon`

    # Add observation numbers

    The test points are in the order in which the data were acquired (consistent with the ANSI/ISA standard). So the observation number is the same as the row number. 

`r code_icon`

```{r}
# observation number is a sequence of integers, from 1 to the number of rows
nrow_in_tidy_data <- nrow(tidy_data)
tidy_data <- tidy_data %>%
	mutate(observ = 1:nrow_in_tidy_data)

# examine
head(tidy_data, n = 3L) 
```

Learn R

- `nrow()` returns the number of rows in the data frame
- `mutate()` creates a new column called `observ`  
- The `:` operator creates a sequence 

## check yourself

Confer with a neighbor.

1. What is the difference between the `observ` variable and the `output_mV` variable?

## simplify the cycle number

`r text_icon`

    # Simplify the cycle number

    The `cycle` data are strings, *cycle_1*, *cycle_2*, etc., from the original column headings. It might be useful to replace these strings with integers, e.g. replacing the string `cycle_1` with the integer `1`. 

`r code_icon`

```{r}
# separate cycle character, convert numerals to integers
tidy_data  <- tidy_data %>%
	separate(cycle, into = c('prefix', 'cycle'), sep = '_', convert = TRUE)

# examine the separated columns
head(tidy_data, n = 3)
```

Learn R

- `separate()` (from  `tidyr`) turns one column of strings into multiple columns of strings 
- `into` assigns the new column names
- `sep` assigns the character separator
- `convert = TRUE` converts the character numerals into integers
- `c()` operator *concatenates* elements to form a vector

`r text_icon`

    The `prefix` column is superfluous. 

`r code_icon`

```{r}
tidy_data <- tidy_data %>%
	select(-prefix)

# examine
glimpse(tidy_data)
```

## final touches

`r text_icon`

    # Final touches 

    The last steps in tidying this data set are to rearrange columns in a logical order and shorten the `test_point` variable name. It's a small enough data set that I can print it. 

`r code_icon`

```{r}
tidy_data <- tidy_data %>%
	select(observ, cycle, test_pt = test_point, input_lb, output_mV)

print(tidy_data)
```

Learn R

- `select()` reorders the columns in the order listed
- `test_pt = test_point` renames the existing column as in `new_name = old_name`

## write to file

`r text_icon`

    Write it to the `data` directory, ready to use for the calibration graph and the regression analysis.   

`r code_icon`

```{r}
write_csv(tidy_data, "data/02_calibr_data-tidy.csv")
```


## check yourself

Your directories should contain these files:

    data\
      |-- 007_wide-data.csv
      |-- 01_calibr_data_active-report.csv
      `-- 02_calibr_data-tidy.csv

    reports\
    
    resources\
      `-- load-cell-setup-786x989px.png 
      
    results\
      `-- 01_calibr_data-wide.csv 
      
    scripts\
      |-- 01_calibr_data-wide.Rmd 
      `-- 02_calibr_data-tidy.Rmd 
      
## push to github

- RStudio Environment pane,`Git` tab  
- Check the `Staged` box for new files and folders 
- `Commit` 
- Type a unique commit message 
- `Commit` 
- `Close` 
- `Push` 
- `Close` 

## check yourself

- Online, navigate to your project repo on GitHub. 
- The new files and folders should be there 

---

[main page](../README.md)  

