---
title: "Reading response"
author: "497 / 597 Reproducible Research"
date: "Fall 2018--19"
output:
  word_document:
    reference_docx: "../resources/word-styles-reference-01.docx"
---

```{r echo=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
opts_chunk$set(echo = FALSE)
library(docxtools)
```


Year / Author ___________________________________________________________

Title of article ___________________________________________________________

- First reading up to a week before the day of discussion.
- Second reading for understanding. Make time to reflect and respond to the prompts.
- Write answers to the following prompts before coming to class. 


In your own words, explain the main thesis of the article.  

```{r fig.height = 1.7}
put_gap()
```

What ideas from the article have direct relevance to your reproducible projects in the course? List as many ideas as you can. 

```{r fig.height = 1.7}
put_gap()
```

What idea from the article did you find most interesting, unexpected, or intriguing? 

