---
output:
  html_document:
    keep_md: yes
  md_document:
    variant: markdown_github
--- 

# preparing a report for transmittal 

```{r 113-01, include = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, results = 'hide', message = FALSE, warning = FALSE, eval = FALSE)
```

```{r 113-02, echo = FALSE, eval = TRUE}
# my tutorial functions for text_icon and code_icon
source("cm/helper_02_icons.R")
options(echo = FALSE) # turns of echo in the console
```


When sharing a report file with someone, the type of file I send them depends on the circumstances. 

- collaborator using RR practices  
- collaborator using Office software 
- client 

## collaborator using RR practices 

If my collaborator is using reproducible practices, I can simply invite them to join my GitHub repo. They can suggest changes, make a pull request that I can review, accept, or reject. 

## collaborator using Office software

If my collaborator is a Word-Excel-PPT user, I send them the `.docx` file. They will typically make revisions using the Word track changes feature, send them back to me, and I have to take the time to incorporate their changes in my "master" script. This adds time to the my workflow, but keeps any document for which I'm responsible reproducible. 


## client 

My usual practice is to make a copy of the `docx` file, rename it, then manually convert it to a PDF file that I can send electronically to the client. When creating the PDF, I add the transmittal date to the file name.

`r code_icon`

```{r}
# make a copy of the docx file
file.copy(from = "reports/06_calibr_report.docx", to = "reports/06_calibr_report_copy.docx"
)
```

Learn R

- `file.copy()` is one of several file management functions in R
- Type `?files` in the Console to browse other file manipulation functions in R
- add `include = FALSE` to the code chunk header to prevent R from printing an output message from this code chunk

On my machine, to create the PDF,

- Manually open the  `06_calibr_report_copy.docx` file
- Print to PDF and edit the file name, e.g., `06_calibr_report_to-client_2016-09-09.pdf`

If later I should edit and knit the report again, the PDF remains unchanged as a  record of the report I sent the client on that date.

[main page](../README.md)








