---
output: github_document
--- 

# project 2 collaboration

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, collapse = FALSE)
```

Part of Project 2 is practicing reproducing someone else's work and then collaborating to improve their work. 


The table below lists each student (the author) and who they will invite to be a collaborator (the reviewer) on their GitHub Project 2 repo. 

## assigned collaborations 

```{r}
library(readxl)
df <- read_excel("manage/project2reviewers.xlsx")
names(df) <- c("Author", "", "Collaborator", "GitHub name")
kable(df)
```




## action items 

The collaborator: 

- Gives their GitHub name to the author 

The author: 

- Receives their collaborator's GitHub name 
- Navigates to their own (the author's) Project 2 repo on GitHub 
- Click on the *Settings* icon 
- Click on the *Collaborators* tab 
- In the *Search by username* box, type the collaborator's GitHub name 
-  *Add collaborator* 

The collaborator: 

- Receives the invitation and has to accept the invitation 




The author: 

- If you haven't already, please invite me (`graphdr`) to also be a collaborator on your Project 2  






--- 
[main page](../README.md)
