---
output: github_document
--- 

# reshaping data from wide to long form

```{r setup, include = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE)
```


Illustrate reshaping a data set from wide form to long form using the built-in `VADeaths` data set. 

```{r}
# load libraries
library(dplyr)
library(tidyr)
```


The `data()` function creates a list of data sets available in R.  

```{r eval=FALSE}
# What data sets are available in my version of R? 
data()
```


One of these in the `VADeaths` data set. Type `?VADeaths` in the Console to see a description of the data set.   

```{r}
# Load the data into your workspace
data(VADeaths)
```


`class()` tells us the data set is a matrix. 

```{r}
class(VADeaths)
```

Let's print the data set. 

```{r}
VADeaths
```

- I see that the age group data is stored in the row names. 
- the column names contain two pieces of information: geographic location (urban or rural) and sex (male or female). 

Before reshaping a data set, plan the final organization. For example, this data set has 4 variables that I'll label `Age_group`, `Geo_region`, `Sex`, and  `Death_rate`. Each variable is a column in the final, tidy data set. Each row is one observation. 

```{r}
# convert the matrix to a data frame
df <- data.frame(VADeaths)
df
```

Copy the data in the row names to a new column. 

```{r}
df <- df %>%
	mutate(Age_group = row.names(df))
df
```

- `mutate()` automatically eliminates the original row names 

```{r}
# Use select() to arrange columns
df <- df %>%
	select(Age_group, Rural.Male:Urban.Female)
df
```

We're ready to use `gather()` to reshape the wide form to long form. 

```{r results = 'hide'}
# gather from wide to long format
df <- df %>%
	gather(Geo_Sex, Death_rate, Rural.Male:Urban.Female)
```

Examine the `gather(Geo_Sex, Death_rate, Rural.Male:Urban.Female)` function one argument at a time 

- `Geo_Sex` is a new column for gathering the *column names* of the original columns labeled  `Rural.Male`, `Rural.Female`, etc. 
- `Death_rate` is a new column for gathering the *column data* of the original columns labeled  `Rural.Male`, `Rural.Female`, etc. 
- `Rural.Male:Urban.Female` tells R the range of columns to be gathered. 

```{r}
# examine the long form data frame after gathering
df
```

- Note that the `Age_group` column from the original data frame is repeated as many times as necessary in the new long form. 

Next, we separate the two pieces of information in the `Geo_Sex` column into separate columns (separate variables). 


```{r}
# separate the two types of data in Geo_Sex to two columns
df <- df %>%
	separate(Geo_Sex, into = c('Geo_region', 'Sex'), sep = '\\.')
```

The character the denotes the separation (in this case) is a period. If you look at the help page for `separate()`, you'll see that the separation character has to be a regular expression., thus we have to "escape" the period by writing `sep = '\.'`. This creates an error too because we have to escape the backslash, hence `sep = '\\.'`. 

```{r}
# examine data frame after separating geography from sex
df
```


We can rearrange the data frame by rows and columns if that helps readability. 


```{r}
# use arrange() to arrange by row 
# use select() to arrange by columns
df <- df %>%
	arrange(Sex, Geo_region, Age_group) %>%
	select(Sex, Geo_region, Age_group, Death_rate)
df
```

Take one last check of the types of data in this data set. 

```{r}
glimpse(df)
```





---

[main page](../README.md)
