---
output: github_document
--- 

# review of tools for preparing data

```{r setup, include = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE)
```


Here, we briefly review the suite of tools we used in the Project 1 tutorials for preparing data for analysis. 

The packages we used are

```{r}
library(dplyr)
library(tidyr)
library(stringr)
library(readr)
library(tibble)
```

These packages are all written by Hadley Wickham, who also happens to be the author of the ggplot2 package and a senior data scientist at RStudio. And quite a nice fellow in person as well. 

If you subscribe to the [RStudio webinar notifications](https://www.rstudio.com/resources/webinars/), you can receive online notices of free webinars put on by Hadley and other folks at RStudio. I recommend it. 
 
## input/output  

- `read_csv()` 
- `write_csv()` 

## examining the data structure  

- `str()`  displays the internal structure of an R object 
- `glimpse()` like `str()` applied to a data frame but it tries to show you as much data as possible 
- `head()` returns the first part of an R object 
- `tail()` returns the last  part of an R object 
- `summary()` returns a statistical summary of an object 
- `colnames()` returns column names 
- `nrow()` returns the number of rows of a data frame 
- `typeof()` tells us what the object is 
- `attributes()`  yields metadata about the object 
- `class()` 


## basic functions in R 

- `max()` and `min()` are R functions that do the obvious 
- `ab()` absolute value 
- `log10()` for base 10, `log()` for base *e* 
- `<=` (less than or equal) and `>=` (greater than or equal) are logical operators that return logical values TRUE or FALSE 


## subsetting data 

- `select()`  keeps only the variables (columns) you list  
- `filter()` keeps the rows listed 
- `[]` brackets, as in `variable_name[indices]` 
- `[[]]` double brackets subsets a list and returns its components 
- `which()` yields the TRUE indices of a logical object 
- `$` extract 


## editing strings  

- `tolower()` strings to lower case 
- `str_replace()` 
- `str_detect()` 
- `str_c()` concatenate strings 
- `separate()` turns one column of strings into multiple columns of strings 


## reshaping data 

- `c()` concatenate elements to create a vector 
- `gather()` collapses multiple columns into two columns: one column of the former column headings (the "keys") and one column of the former column entries (the "values"). All other columns are duplicated (copy/paste downward in chunks) as needed. 
- `mutate()` creates a new column in a data frame (can also be used to overwrite an existing column) 
- `unname()`  removes the names from named numbers 
- `frame_data()` create a data frame row by row 
- `data.frame()` convert object to data frame 
- `arrange()` arrange by row entries 
- `select()` arrange by column names 


## display results  

- `kable()` 
- `print()` write an object in the Console 


## control of code execution 

- `stopifnot()` stops the code execution and produces an error message if any argument returns FALSE

---

[main page](../README.md)
