---
output: github_document
--- 

# data preparation

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = FALSE)
```

```{r echo = FALSE}
# functions for text_icon and code_icon
source("cm/helper_02_icons.R")
```


## overview of chapter 7 

This tutorial follows material in Chapter 7 in the text. Topics include: 

- Getting a handle on your data
- Reshaping your data 
- Renaming variables 
- Ordering variables 
- Subsetting 
- Recoding 
- Creating new variables 
- Factors 
- Changing data types 
- Binding 
- Merging 
- Duplicates 
- Complete or incomplete cases 
- Missing values 
- NAs 



## start a new script in practiceR 

In `practiceR/scripts/`, start a new Rmd file called `chapter7.Rmd`. Write the code chunks shown below with as much of the prose as yo9u like  to explain the work. 

`r code_icon` Start with the usual code chunk. 

```
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE)
```




## load a new data set for the exercises (section 6.3.4)

Install the `WDI` package (World Bank Development Indicators) 

In RStudio: *Package > Install > WDI* 

`r code_icon` 

```{r}
# load the WDI package 
library(WDI)
```

The `WDI` package includes its own search function.  

`r code_icon` 

```{r eval = FALSE}
# search for the data on consumption of fertilizer
WDIsearch("fertilizer consumption") 
```

The output of this function is shown below. Recall that code output is shown using a double hashtag `##`. It is shown here so you can check your work...don't copy and paste output like this into your script. 

```{r echo=FALSE}
# search for the data on consumption of fertilizer
WDIsearch("fertilizer consumption")
```

The subset we'll use from this data is the `AG.CON.FERT.ZS`. 

`r code_icon`

```{r}
FertConsumpData <- WDI(indicator = "AG.CON.FERT.ZS")
```


Check you Environment pane. You should see the new object **FertConsumpData**. 

Our author uses CamelCase for variable names. My current practice generally is to use all lowercase letters separated by underscores, for example, **fert_consump_data**. But, because I'm following the text example, I'll use the text convention.  

## getting a handle on your data (7.1.1)

We covered most of the following functions in the calibration report tutorials. Use them to study your data before starting any data tidying 

- `str()` 
- `glimpse()` 
- `dim()` 
- `head()` and `tail()` 
- `summary()` 
- `rownames()` and `colnames()` 
- `nrow()` and `ncol()` 
- `typeof()` 
- `attributes()` 
- `class()` 

The data we downloaded looks like this: 

`r code_icon` 

```{r}
head(FertConsumpData)
```

Looks like we have 4 variables. What data structure do we have? 

`r code_icon` 

```{r}
library(dplyr)
glimpse(FertConsumpData)
```

- The number of observations is the number of rows. 
- The number of variables is the number of columns. 
- This data set appears to be tidy already. 

`r code_icon` 

```{r}
dim(FertConsumpData)
```

- `dim()` returns the dimensions of the object 
- Because **FertConsumpData** is a 2-dimensional array, we could also use `nrow()` and `ncol`

`r code_icon` 

```{r}
nrow(FertConsumpData)
ncol(FertConsumpData)
colnames(FertConsumpData)
```

Examine basic statistics of the 4 columns 

`r code_icon` 

```{r}
summary(FertConsumpData)
class(FertConsumpData)
```

- From the summary, we see that we have missing values (NAs)
- The object is a data frame (which we like)  


--- 
[main page](../README.md)
