---
output: github_document
--- 

# graph tour overview  

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = FALSE, fig.keep = 'high')
```

```{r echo = FALSE}
# functions for text_icon and code_icon
source("cm/helper_02_icons.R")
```

There are currently four graphical systems available in R:  

- `base` graphics, by Ross Ihaka, is included in every R installation 
- `grid` graphics, by Paul Murrell, provides low-level access to graphic primitives 
- `lattice`, by Deepayan Sarkar, excels at displaying multivariate data 
- `ggplot2`, by Hadley Wickham, is based on a comprehensive graphical "grammar" 

I've used `lattice` for years now, but have recently started learning `ggplot2`. In my opinion, it's the package to learn; all the cool kids at the R conference are using it.  Read more about `ggplot2` in section 10.4 of the text. 


As you work through the tour, create a new script for each graph type in your 
`practiceR/scripts/` directory. I hope to be able to show you examples of these commonly encountered graph types:  

- Cleveland dot plot 
- histogram 
- box and whisker plot 
- scatter plot 
- time series 
- multiway plot 

 
--- 

[main page](../README.md)
