---
output: github_document
--- 



# introduction to factors 

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, fig.keep = 'high')
```

```{r echo = FALSE}
# functions for text_icon and code_icon
source("cm/helper_02_icons.R")
```

As a reminder, the commonly encountered vectors types in R:  

- double (numeric double precision) 
- integer (numeric integers)
- complex (numbers), e.g., `z <- 2 + 5i` 
- character (strings) 
- factor `<-` today's topic 



## categorical data 

*Categorical variables* in a data set  describe qualitative properties. The range of possible values (or levels) a categorical variable can assume is finite. Two examples of categorical data and their levels: 

- Category "cardinal directions" has 4 levels (north, west, south, east) 
- Category "gender identity" has [58 levels](http://abcnews.go.com/blogs/headlines/2014/02/heres-a-list-of-58-gender-options-for-facebook-users/) (agender, androgyne, androgynous, bigender, cis, cisgender, cis female, cis male. etc.). Other sources list even more levels.  

Some categorical data have an intrinsic order (*ordinal* data) and others do not (*nominal* data). For example, the data values "low", "medium", and "high" have an intrinsic order while the data values "north", "west", "south", and "east" do not.  

*Factors* are a data type in R specialized for categorical data, ordered or  unordered.     



## start a new script in practiceR 

- In `practiceR/scripts/`, start a new Rmd file called `factor_play.Rmd`. 
- Copy the code chunks in this tutorial into `factor_play.Rmd` and add as much prose as you like to explain the work. 

`r code_icon` 
```{r}
library(dplyr)
library(readr)
```

The standard R installation includes US "state" data sets, with all data arranged according to alphabetical order of the state names. (Learn more by typing `?state`  in your Console.) There are 7 data sets, named:  

```
state.abb 
state.area 
state.center 
state.division 
state.name 
state.region 
state.x77 
````
Let's take a look at several variables, one numerical (area) and three that are categorical (abbreviation, name, and region): 

`r code_icon`
```{r}
# examine the vector of abbreviations
glimpse(state.abb)

# examine the vector of names  
glimpse(state.name)

# examine the vector of areas  
glimpse(state.area)

# examine the vector of regions 
glimpse(state.region)
```

Of the categorical variables, two are encoded as characters (abbreviation and state) and one is encoded as a factor (region). 


Let's collect these 4 vectors in a new data frame, renaming the variables to eliminate the "dot" in the variable names and writing the data to file for later use. 

`r code_icon`
```{r}
# create a new data frame
state_df <- data_frame(abbr = state.abb, 
		name   = state.name, 
		area   = state.area, 
		region = state.region)

# write the df to file for later use 
write_csv(state_df, "data/cm034_state.csv")
```




## strings or factors

We saw above that categorical data in R can be represented as strings or factors. The data frame we created above retained the data types of the original vectors: 

```{r}
# examine the new data frame 
glimpse(state_df)
```


- `abbr <chr>` is a vector of character strings 
- `name <chr>` is a vector of character strings 
- `area <dbl>` is a double-precision number 
- `region <fctr>` is a factor 

You will encounter categorical data in both forms, characters and factors, and should recognize the attributes of both: 

- Character strings are suitable for unordered (nominal) data. 
- Factors are suitable for either type, ordered or unordered.  
- Factor levels are *viewed* as character strings but are *encoded* as integer values $1...k$. 
- Integer encoding (hidden from casual view) assigns an implicit order, even for unordered values.  

*Integer encoding* is the aspect of factors that is both useful and sometimes puzzling if it catches you unaware. Hence this tutorial. 



## factors are encoded as integers

Let's look more closely at the attributes of the factor. I'll start by extracting part of the region column as a stand-alone vector.  

`r code_icon`  
```{r}
# extract the region column, first fifteen rows  
first_15 <- state_df$region[1:15]

# confirm region is a factor
class(first_15)
 
# factors have levels 
levels(first_15)
```

The `levels` attribute tells us the vector has 4 distinct values, in the order shown:  

- level 1 is "`r levels(first_15)[1]`" 
- level 2 is "`r levels(first_15)[2]`" 
- level 3 is "`r levels(first_15)[3]`" 
- level 4 is "`r levels(first_15)[4]`" 

Printing out the values, the data appear to be characters: 

`r code_icon` 
```{r}
# the region data appear to be characters  
first_15
```

But `typeof()` tells us that this vector is composed of integers:   

`r code_icon` 
```{r}
# the region data are of type "integer" 
typeof(first_15)
```

We can see the integers themselves by removing the "class" attribute, 

`r code_icon` 
```{r}
# the values are actually encoded as integers 
unclass(first_15)[1:15]
```

The integers represent the factor levels: 


- `r unclass(first_15)[1]` is `r first_15[1]` (`r state_df$name[1]`) 
- `r unclass(first_15)[2]` is `r first_15[2]` (`r state_df$name[2]`) 
- `r unclass(first_15)[3]` is `r first_15[3]` (`r state_df$name[3]`) 
- `r unclass(first_15)[4]` is `r first_15[4]` (`r state_df$name[4]`) 
- `r unclass(first_15)[5]` is `r first_15[5]` (`r state_df$name[5]`) 
- `r unclass(first_15)[6]` is `r first_15[6]` (`r state_df$name[6]`) 
- `r unclass(first_15)[7]` is `r first_15[7]` (`r state_df$name[7]`) 
- etc. 

This example illustrates that factors in R always have an implicit order, even for unordered (nominal) data, because factors are encoded as integer sequence $1 ... k$. 




## summary

- Categorical data can be represented in R as strings or factors. 
- Categorical data can be intrinsically ordered or unordered. 
- Elements of a factor vector are its "levels". 
- The number of levels $k$ is the number of unique values in the vector. 
- Levels are encoded in R as integers representing the level numbers, $1...k$. 





--- 
[main page](../README.md)
