---
output: github_document
--- 

# histogram  

```{r cm036-01, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, fig.keep = 'high')
```

```{r cm036-02, echo = FALSE}
# functions for text_icon and code_icon
source("cm/helper_02_icons.R")
```

## getting started 

We're going to use data from the 2011 World Speed Skiing Championships. The data set is called `SpeedSki` from the `GDAdata` package. GDA stands for "Graphical Data Analysis"; the package accompanies the book by Unwin (2015). 

- First, install the `GDAdata` package (In RStudio, *Packages > Install > GDAdata*) 
- You can then find out all about the GDAdata package by using the RStudio Package window, scroll down until you find `GDAdata`, and click on its name. A help window should appear.  
- Create `histogram.Rmd` in your `practiceR/scripts/` directory. Write the code chunks in the tutorial with as much of the prose as you like  to explain the work. 

## data 

Load the package and examine the `SpeedSki` data frame. The data frame is already tidy. 

`r code_icon`
```{r cm036-03}
# load the package that has the data we want 
library(GDAdata)

# load the data set 
data(SpeedSki)

# examine the data
library(dplyr)
glimpse(SpeedSki)
```

Variable **Speed** is a number (km/hr) while variables **Event** and **Sex** are factors (useful for conditioning). 

First graph is a histogram of the counts of all speeds. 
 
 `r code_icon`
```{r cm036-04}
library(ggplot2)
p <- ggplot(SpeedSki, aes(x = Speed)) + 
	geom_histogram()

p # to print the graph
```

We can play around with the bin-width to adjust the resolution of the histogram. A bin width of 2 km/hr gives a useful distribution. You can try others. In this chunk, I also set my own x-scale limits. 

`r code_icon` Edit the earlier code chunk 
```{r cm036-05}
# histogram bin width
set_bin <- 2

# plot 
p <- ggplot(SpeedSki, aes(x = Speed)) + 
	geom_histogram(binwidth = set_bin) + 
	xlim(160, 220)

p # to print the graph
```

Let's condition the data by Sex. The conditioning is assigned using the `fill = Sex` argument of the `aes()` specification. Conditioning automatically assigns a color palette and inserts a legend. 

`r code_icon` Edit the earlier code chunk 
```{r cm036-06}
p <- ggplot(SpeedSki, aes(x = Speed, fill = Sex)) + 
	geom_histogram(binwidth = set_bin) + 
	xlim(160, 220)

p # to print the graph
```

Facets (multiple panels) are created using `facet_wrap()`. By disaggregating the data by Sex, we eliminate any overlap in the two groups of data. The  `ncol= 1` argument aligns the two panels in a column. 

`r code_icon`
```{r cm036-07}
# the facet can be added as a layer 
p <- p + facet_wrap(~ Sex, ncol = 1)

p # to print the graph
```

The panels are labeled so we can dispense with the legend. 

`r code_icon`
```{r cm036-08}
# omit the legend
p <- p + theme(legend.position = "none")

p # to print the graph
```

As Unwin notes in his discussion of these data, 

> As well as emphasizing that there were many more competitors in the men's competition than in the women's, the plots show that the fastest person was a man and that a woman was the slowest. What is surprising (and more interesting) is that the fastest woman was almost as fast as the fastest men and that there were two distinct groups of women, the fast ones and the slow ones.

Conditioning the data further by Sex and Event helps explain these results. 

`r code_icon`
```{r cm036-09}
p <- p + facet_grid(Sex ~ Event)

p
```

A much clearer picture emerges. 

- Speed One is the fastest event (requiring special equipment). 
- In the Speed One competition, women were faster than the majority of men.  
- No women took part in the Downhill.  
- There is not much variation among the Juniors. 
- The reason for the two female groups is now clear: They took part in two quite different events. 

Some final tweaks to make the graph more presentable. 

`r code_icon`
```{r cm036-10}
# outline the bins
p <- p +
	geom_histogram(binwidth = set_bin, color = "white")

# edit the x and y scale labels 
p <- p +
	  xlab("Speed (km/hr)") + 
    ylab("")

# edit panel background color to a slightly darker shade of gray 
p <- p +
	theme(panel.background = element_rect(fill = "gray90", colour = "gray90", size = 0.5, linetype = "solid"))

# create some custom colors 
library(RColorBrewer)
BrBG  <- brewer.pal(5, "BrBG")
medBr <- BrBG[1]
medBG <- BrBG[5]

# re-color the bins
p <- p +
	scale_fill_manual(values = c(medBr, medBG)) 

p
```



## references


1. Antony Unwin (2015) *Graphical Data Analysis with R*, CRC Press. 




--- 
[main page](../README.md)
