---
output: github_document
--- 

# project 3 getting started 

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, collapse = FALSE)
```




## assigned pairs 

```{r}
library(readxl)
df <- read_excel("manage/project3collab.xlsx")
names(df) <- c("Project", "Partner", "GitHub", "Partner", "GitHub")
kable(df)
```


### action items 

Alice and Carlos, the two partners, meet to agree on: 

- Who will be Alice and set up the initial repo 
- What the initial directory structure will be 

Alice: 

- Creates a private project 3 repo 
- Clones the repo locally and starts setting up a directory structure 
- Invites the other team member as collaborator 
- Invites me (graphdr) to be a collaborator 

Carlos: 

- Accepts the invitation 
- Clones the remote repo locally 

Alice and Carlos:  

- Read the full [abstract online](cm031_project-3-descriptions.md) and the associated links. 
- Jointly draft the README.md for the repo. You may incorporate elements of my  abstracts to start. 
- Begin planning your tasks for obtaining the data and starting the first script 



## collaborators for review 

Like we did in Project 2, each pair of you invites another pair to be your collaborators for the purposes of checking the reproducibility of your project. 

```{r echo=FALSE}
library(dplyr)
library(stringr)
df <- read_excel("manage/project3reviewers.xlsx") %>%
	mutate(github_name_1 = str_c(github_name_1, ", ", github_name_2)) %>%
	select(authors, invites_reviewers, github_name_1)
names(df) <- c("Authors", "Invite reviewers", "GitHub names")
kable(df)
```

### action items 

Authors: 

- One of you invites the two reviewers (listed above) as collaborative reviewers on your project 

Reviewers: 

- Accept the invitation from the Author team 
- Clone the invited project locally so you can do your review work as the term progresses 



--- 
[main page](../README.md)
