---
output: github_document
--- 

# line graph  

```{r cm045-01, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, fig.keep = 'high')
```

```{r cm045-02, echo = FALSE}
# functions for text_icon and code_icon
source("cm/helper_02_icons.R")
```

## getting started 

We're going to use the car production data from Figure 8.13 in Robbins. I've saved it in a csv file you can download from the course repo. 

- Download the data file `fig-08-13_data-car-production.csv` and save it in your `practiceR/data/` directory  
- Create `line-graph.Rmd` in your `practiceR/scripts/` directory. Write the code chunks in the tutorial with as much of the prose as you like  to explain the work. 

## data 

The data frame is already tidy. 

```{r cm045-03}
library(readr)
library(dplyr)
df <- read_csv('data/fig-08-13_data-car-production.csv')
glimpse(df)
```

For conditioning, we'd like the country variables to be a factor instead of a character. 

```{r cm045-04}
df <- df %>%
	mutate(country = factor(country))
glimpse(df)
```

## graph

Ignoring country for the moment, let's graph production by year. We'll use a `geom_line()` and a `geom_point` to shows the individual data markers and to connect the dots. 

```{r cm045-05}
library(ggplot2)
ggplot(df, aes(x = year, y = production)) +
  geom_line() +
  geom_point()
```

Well, that's a mess. 

Do it again, but add the `group` argument to the `aes()` function so the data are grouped by country. This grouping is why we converted country from characters to a factor. 

```{r cm045-06}
ggplot(df, aes(x = year, y = production, group = country)) +
  geom_line() +
  geom_point()
```

That's better, but we don't know which country is which. For that, we add the `color` argument to the `aes()` function where we want to group and color both by country. 

```{r cm045-07}
ggplot(df, aes(x = year, y = production, group = country, color = country)) +
  geom_line() +
  geom_point()
```

Notice that adding the `color` argument automatically added the legend. 

Sometimes, multiple lines on a graph are so close in value, it is difficult to see the individual trends. In such a case, we want to use a trellis (or small multiple) design, invoked by the `facet_grid()` function. 

Because we are switching to facets, we delete the group and color by country we used before and add the `facet_grid()` function.  

```{r cm045-08}
ggplot(df, aes(x = year, y = production)) +
  geom_line() +
  geom_point() +
  facet_grid(.~ country)
```

## panel order 

For quantitative data, the panels should be ordered, usually in order of increasing values, e.g., the production data. 

First, let's look at the existing order of the levels of the country column (a factor). 

```{r cm045-09}
# confirm that the country variable is a factor
class(df$country)

# check the order of the factor levels 
levels(df$country)
```

Alphabetical, as expected. The graph panels from left to right are also in this order. 

To change the order of the graph panels, we have to change the order of the factor levels. In a previous example using factors, we simply typed  the levels in the order we wanted them using the `levels` argument of the `factor()` function. 

However, in this example we want to automate the reordering of the factor levels based on the data in the production column. 

To start, let's print the data frame. 

```{r cm045-10}
# take a look at the data frame
df
```

Then create a vector of indices ordered by the production numbers. 

```{r}
# determine the order of the production numbers 
production_order <- order(df$production)

# ordering produces a vector of integers (indices)
production_order 
```

By default, `order()` places the elements of its argument in ascending order.  

- The first element in `production_order` is `r production_order[1]`. Looking at the data frame, we see that the row with the lowest production number is `r df[production_order[1], ]`, in row `r production_order[1]`.  
- The last element in `production_order` is `r production_order[20]`. Looking at the data frame, we see that the row with the highest  production number is `r df[production_order[20], ]`, in row `r production_order[20]`.  

Thus `production_order` is essentially a vector of row numbers (or indices) that orders the rows of the data frame in increasing order of production.  We can show that. 

```{r}
# reorder the rows of the data frame and print
df[production_order, ]
```

Next we use `production_order` to reorder the country level encoding.  Recall, the country variable is a factor, encoded by integers $1:`r length(unique(levels(df$country)))`$. 

```{r}
# reorder the country panels by increasing magnitude of production 
new_levels <- unique(df$country[production_order])

# show the new ordering of levels 
new_levels
```

You can see that the levels are no longer in alphabetical order. Instead they are in order of increasing production. 

Finally, we use the `levels` argument of the `factor()` function to assign the new levels in the order we've found. 

```{r}
df$country <- factor(df$country, levels = new_levels)
```

Important to note that the reordering of the levels has no effect on the ordering of the rows of the data frame. All we've done is reorder the integers that encode the factor variable. 

Finally, the code to this point has been presented in small chunks for understanding, but we can actually do the same work with one line: 

```{r}
# reorder country levels by order of increasing production 
df$country <- factor(df$country, 
	  levels = unique(df$country[production_order])
	  )
```

Create the graph again using the same code as before. The result shows that from left top right, the panels are in increasing order of production. 

```{r cm045-12}
p <- ggplot(df, aes(x = year, y = production)) +
  geom_line() +
  geom_point() +
  facet_grid(.~ country)  
p
```


## edit for publication 

There is a bit of overprinting in the time scale, so we can edit the year labels to show only every other year. 

```{r cm045-13}
p <- p + scale_x_continuous(labels = c(1977, "", 1979, ""))
p
```

Next, axis labels and a title. 

```{r cm045-14}
p <- p + labs(x = "Year", y = "Percent") +
  ggtitle("Fig 8.13 Car production, 1977-80")
p
```


Of course, you could continue to edit for an appearance that you find suitable for your audience. 
  
  

## references


Figure 8.13 is from Robbins. The `ggplot2` code is adapted from Zhao & Bryan.  

1. Naomi Robbins (2013) [*Creating More Effective Graphs*](http://www.nbr-graphs.com/resources/recommended-books/), Chart House.     
2. Joanna Zhao and Jenny Bryan,  [r-graph-catalog](https://github.com/jennybc/r-graph-catalog), providing ggplot2 scripts for nearly all of the graphs in Robbins' book.    
3. Winston Chang, [Facets (ggplot2)](http://www.cookbook-r.com/Graphs/Facets_(ggplot2)), [Cookbook for R](http://www.cookbook-r.com/).  





--- 
[main page](../README.md)
