---
output: github_document
--- 

```{r echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, results = 'hide')
```

```{r echo = FALSE}
source("cm/helper_02_icons.R")
```

# Rmd to docx: numbering tables and figures

As Norbert K&ouml;hler says, "R Markdown has no native method to number and reference table and figure captions". The `captioner` packages gives us a method  for auto-numbering figures and tables and automatically updating the in-text reference numbers  as well. 

This tutorial is adapted from Norbert K&ouml;hler's [blog post](http://datascienceplus.com/r-markdown-how-to-number-and-reference-tables/) plus Andrew Dolman's [comments](https://onedrive.live.com/?authkey=%21AApYWBGHNR06Y5I&cid=FCFCD3FD042AF7F1&id=FCFCD3FD042AF7F1%2118306&parId=FCFCD3FD042AF7F1%21322&action=locate) on that post. 


When complete, your output document should be a 2-page docx file with two tables and two figures with automatically-generated table and figure numbers. 

## install 

- Install the `captioner` package 
- Install the `devtools` package 
- In your Console: `devtools::install_github('yihui/knitr', build_vignettes = TRUE)` 

The purpose of last line above is to install the latest development version of `knitr` so we can take advantage of the knitr option `row.names = FALSE` when using `kable()` to print a table. Plus we get recent bug fixes. 

## getting started 

- Launch your `practiceR` project 
- Open a new Rmd file. Mine is called `test-captioner.Rmd`. 
- Both `word_document` and `html_document` can be used in the YAML header.  
- For `pdf_document`, see the references for additional lines needed. 

The YAML header. 

```
---
title: "Testing *captioner*"
output:
  word_document:
    fig_caption: yes
---
```

The usual setup. 

```{r}
# the usual setup
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE)
```

Load packages. 

```{r}
# packages  
library(captioner)
library(ggplot2)
library(splines)
```


## assign prefixes

Use `captioner()` to set the prefixes we want to use for our tables and figures. Here, I'm using the full words. Some folks prefer to use abbreviations,  "Fig." and "Tab.". 

```{r results = 'hide'}
# assign the prefixes
auto_tab_ID <- captioner(prefix = "Table")
auto_fig_ID <- captioner(prefix = "Figure")
```

The function names on the left, `auto_tab_ID` and `auto_fig_ID`, are arbitrary. 

## create a table caption  

Begin the report. Start with a level 1 heading. 

    # Tables 

We create *table captions* using `auto_tab_ID()`. The `name` argument is a unique string to identify the table prefix and `caption` string.

```{r results = 'hide'}
# assign the table prefix, number, and caption 
auto_tab_ID(name = "a_table", caption = "A descriptive caption for the table.")
```



The table number (in this case, Table 1) is determined by the order of appearance of the `auto_tab_ID()` function in the script. This is its first use, so the table number is "1". 


## refer to the table inline  

In a typical technical report, the paragraph just before a table includes an inline reference to the table. For example:  

> An excerpt of data from our fabulous experiment is shown in `r auto_tab_ID("a_table", display = "cite")`. 

To construct this sentence, we an inline code chunk and the function  `auto_tab_ID()` with an added argument `display = "cite"` to print the table number but not the caption. 

<pre><code>An excerpt of data from our fabulous experiment is shown in 
<code>`</code>r auto_tab_ID("a_table", display = "cite")<code>`</code>. 
</code></pre>

## kable() the table 

Keeping things simple, I'll use `knitr::kable()` to print data from the `cars` data set. 

The `kable()` function has a `caption` argument we use to invoke the `auto_tab_ID()` function.  

```{r}
# get some data
df <- head(cars, n = 5L)

# print the table with caption
kable(df, caption = auto_tab_ID("a_table"), row.names = FALSE)
```

I haven't printed the table here, but it should appear in your docx output. 

In docx output, table captions are formatted with the *Table Caption* style. The style can be edited using methods described in [Controlling Word Styles](https://github.com/DSR-RHIT/me497-RR/blob/master/cm/cm041_word-styles.md).

## make a second table 

I'm going to create another table to demonstrate that the next table is indeed  assigned the number "2". 

```{r results = 'hide'}
# create a caption for a second table 
auto_tab_ID("another_table", "A boring table.")
```

Having created another caption, I can make an inline reference to the next table.

> More amazing results are shown in `r auto_tab_ID("another_table", display = "cite")`. 

Created using this Rmd script: 

<pre><code>More amazing results are shown in 
<code>`</code>r auto_tab_ID("another_table", display = "cite")<code>`</code>. 
</code></pre>

New table: 

```{r}
# get some data
df <- tail(cars, n = 5L)

# print the table with caption
kable(df, caption = auto_tab_ID("another_table"), row.names = FALSE)
```

All tables are automatically renumbered if tables are added or deleted. 



## create a figure caption  

Create another level 1 heading. 

    # Figures 

Create my first figure caption using  `auto_fig_ID()`. 

```{r results = 'hide'}
auto_fig_ID(name = "one_figure", caption = "An auto-numbered figure caption.")
```

The figure number (in this case, Figure 1) is determined by the order of appearance of the `auto_fig_ID()` function in the script. This is its first use, so the figure number is "1". 


## refer to the figure inline 

Inline references are typical. 

> More amazing results are shown in `r auto_fig_ID("one_figure", display = "cite")`. 

Created using this Rmd script: 

<pre><code>More amazing results are shown in 
<code>`</code>r auto_fig_ID("one_figure", display = "cite")<code>`</code>. 
</code></pre>



## make a figure 

In contrast to table captions, we create  *figure captions* using the `fig.cap` argument inside `{r }` header that starts a code chunk  


For example, create a code chunk with these arguments:

<pre><code>```{r fig.cap = auto_fig_ID("one_figure"), fig.height = 3}

<code>```</code></code></pre>

and with this graph code:  
 
```{r fig.cap = auto_fig_ID("one_figure"), fig.height = 3, eval = FALSE}
ggplot(data = cars, aes(x = speed, y = dist)) + 
	geom_point() + 
	geom_smooth(method = "lm", formula = y ~ bs(x, 3), se = FALSE)
```

Just for fun, I've included a spline fit: 

- `bs()` is a polynomial B-spline from the `splines` package 
- For a complete list of spline functions, use `library(help = "splines")` 

## make a second figure  

Demonstrate the next figure is auto-numbered "2". 

New caption: 

```{r results = 'hide'}
# create a caption for a second figure 
auto_fig_ID("another_figure", "Same data with a linear fit.")
```

New inline reference: 

> If we do the same graph with a linear fit, we get
`r auto_fig_ID("another_figure", display = "cite")`. 

Created using this Rmd script: 

<pre><code>If we do the same graph with a linear fit, we get 
<code>`</code>r auto_fig_ID("another_figure", display = "cite")<code>`</code>. 
</code></pre> 

New code chunk:

<pre><code>```{r fig.cap = auto_fig_ID("another_figure"), fig.height = 3}

<code>```</code></code></pre>

with this graph code, using a linear regression instead of a spline. 

```{r fig.cap = auto_fig_ID("another_figure"), fig.height = 3, eval = FALSE}
ggplot(data = cars, aes(x = speed, y = dist)) + 
	geom_point() + 
	geom_smooth(method = "lm", se = FALSE)
```

That's all folks! 


## references

1. Norbert K&ouml;hler (2016-10-22) [R Markdown: How to number and reference tables](http://datascienceplus.com/r-markdown-how-to-number-and-reference-tables/) 
2. Andrew Dolman (2016) [Testing captioner](https://onedrive.live.com/?authkey=%21AApYWBGHNR06Y5I&cid=FCFCD3FD042AF7F1&id=FCFCD3FD042AF7F1%2118306&parId=FCFCD3FD042AF7F1%21322&action=locate) from the comments section of Norbert's blog post. 


--- 
[main page](../README.md)
