---
output: github_document
--- 

# run all scripts

```{r echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE)
```

When you want to run an entire project's worth of scripts from beginning to end, the "proper" approach is to use *makefiles*. We haven't discussed makefiles in class, but the text gives a good introduction if you are interested. 

Makefiles save time by only executing those the scripts that have changed since the last time the makefile was run. 

The approach I show you here runs all the scripts every time---so it takes more time than makefiles. For most of our projects, however, the time penalty shouldn't be too annoying. 

## start 
 
This code will knit all your Rmd scripts in sequence. The assumptions are: 

- we want to compile all the scripts in the `scripts/` directory 
- the Rmd scripts are named to render in order, e.g., filenames start with 01, 02, 03, etc. 

Start a new Rmd script `render_scripts.Rmd` and save it to your `scripts/` directory. 

Start with the usual knitr initialization.

```
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE)
```

We're need the `rmarkdown` package to knit each script.  

```{r}
# package
library(rmarkdown)
```

## find the Rmd scripts 

Name the path to the directory that contains the scripts we want to knit.   

```{r}
#  identify the path to the directory we want 
path_to_Rmds <- "scripts"
```

Use `list.files()` to find all the Rmd scripts in that directory and save them in `Rmd_scripts`, a vector of character strings. 

```{r}
# find all files that end in .Rmd
Rmd_scripts <- list.files(
	path = path_to_Rmds
	, pattern = "\\.Rmd$"
	, full.names = TRUE
)
```


## knit the Rmd scripts 

For each filename in `Rmd_scripts`, apply the `render()` function one file at a time. The "one-file-at-a-time" process is managed by the `sapply()` function. 

```{r eval = FALSE}
# render each script
sapply(Rmd_scripts, function(x) render(x))
```


## run it 

Run the script and all your Rmds should be rendered in order. 


--- 
[main page](../README.md)
