---
output: github_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)
opts_chunk$set(collapse=TRUE)
```

# graphs: gettings started

R for Data Science online [Sections 3.1 and 3.2](http://r4ds.had.co.nz/data-visualisation.html#introduction-1)

## section 3.1.1

```{r}
library(tidyverse)
```

## section 3.2.1

`mpg` is a data set in the tidyverse.

```{r}
mpg
```


## section 3.2.2


```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

## exercises 3.2.4  

Do the exercises on your own for practice with the concepts

## section 3.3

With `class` mapped to the `color` aesthetic. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

With `class` mapped to the `size` aesthetic---produces a warning.  

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

With `class` mapped to the `alpha` aesthetic---produces a warning. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

With `class` mapped to the `shape` aesthetic---produces a warning. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Set the aesthetic properties of the geom manually, outside of `aes()`. 

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue")
```

## exercises 3.3.1 

Do the exercises on your own for practice with the concepts
