---
output: github_document
bibliography: "../resources/bib-etc/references-rr.bib"
csl: "../resources/styles/apa.csl"
link-citations: yes
--- 

# data tools outline

```{r setup, include = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE)
```

Here, we briefly review the suite of tools we use for preparing data for analysis.

```{r}
library(tidyverse)
library(stringr)
```


If you subscribe to the [RStudio webinar notifications](https://www.rstudio.com/resources/webinars/), you can receive online notices of free webinars put on by Hadley Wickham and other folks at RStudio. I recommend it. 
 
## read and write 

- `read_csv()`  and  `write_csv()` 
- `read_lines()` and `write_lines()`
- `read_table()` 
- `read_excel()`
- readr argument `convert = TRUE` 

## examining the data structure  

- `str()`  displays the internal structure of an R object 
- `glimpse()` like `str()` applied to a data frame but it tries to show you as much data as possible 
- `head()` returns the first part of an R object 
- `tail()` returns the last  part of an R object 
- `summary()` returns a statistical summary of an object 
- `names()` returns column names 
- `nrow()` returns the number of rows of a data frame 
- `typeof()` tells us what the object is 
- `attributes()`  yields metadata about the object 
- `kable()` print a knitr table 
- `print()` write an object in the Console 
- `class()` 
- `unique()` 
- `levels()` 


## basic functions in R 

- `abs()` absolute value 
- `log10()` for base 10 
- `log()` for base *e* 
- `<=` (less than or equal) and `>=` (greater than or equal) are logical operators that return logical values TRUE or FALSE 
- `c()` concatenate elements to create a vector  
- `unname()`  removes the names from named numbers 
- `unlist()` 
- `mean()`, `sum()`, `median()`, `sd()`, `max()`, `min()`, `lm()` 


## subsetting data 

- `select()`  keeps only the variables (columns) you list  
- `filter()` keeps the rows listed 
- `[]` brackets, as in `variable_name[indices]` 
- `[[]]` double brackets subsets a list and returns its components 
- `which()` yields the TRUE indices of a logical object 
- `$` extract 

## tidying and transforming data 

- `separate()` 
- `unite()` 
- `gather()` 
- `spread()` 
- `arrange()`  
- `mutate()` 
- `group_by()` and `summarize()` 
- `n()` 
- `as.integer()`, `as.double()`, `as.character()` 


## editing strings  

- `tolower()` strings to lower case 
- `str_replace()` 
- `str_detect()` 
- `str_c()` concatenate strings 
- `separate()` turns one column of strings into multiple columns of strings 
- other stringr package 


## control of code execution 

- `stopifnot()` stops the code execution and produces an error message if any argument returns FALSE

