---
output: github_document
bibliography: "../resources/bib-etc/references-rr.bib"
csl: "../resources/styles/apa.csl"
link-citations: yes
--- 



```{r}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo=TRUE, collapse=TRUE)
```


# reading messy spreadsheet data

I tried the jailbreakr package, but it isn't on CRAN and the repo doesn't have any useful vignettes. There is a blog post though at  https://www.brodrigues.co/blog/2017-02-17-how_to_use_jailbreakr/ 

```{r}
# install just once
# devtools::install_github(c("hadley/xml2",
#                            "rsheets/linen",
#                            "rsheets/cellranger",
#                            "rsheets/rexcel",
#                            "rsheets/jailbreakr"))
```


Following the instructions at the blog, 

```{r}
library("rexcel")
library("jailbreakr")

path_to_data <- "data/us_income.xlsx"
my_sheet     <- rexcel_read(path_to_data, sheet = "us_income")
tables       <- split_sheet(my_sheet)
```

Not sure what that all is. But keep going 

```{r}
library(purrr)
list_of_data <-  map(tables, (function(x) (x$values())))
# map(list_of_data, head)
```

The variable `list_of_data` contains the discrete chunks of the messy spreadsheet. 

```{r}
j <- 1
dataset_j <- list_of_data[[j]]
dataset_j


j <- 2
dataset_j <- list_of_data[[j]]
dataset_j


j <- 3
dataset_j <- list_of_data[[j]]
dataset_j

j <- 4
dataset_j <- list_of_data[[j]]
dataset_j


j <- 5
dataset_j <- list_of_data[[j]]
dataset_j

j <- 6
dataset_j <- list_of_data[[j]]
dataset_j


j <- 7
dataset_j <- list_of_data[[j]]
dataset_j
```


Stop here and compare an approach using the tidyverse. At this point I can pull a matrix for each state. 


## using the tidyverse

```{r}
library(tidyverse)
```

```{r}
rawlines <- read_lines("data/us_income.csv")
head(rawlines)
```

Find the row with Alabama

```{r}
str(rawlines)

grep("Alabama", rawlines, ignore.case = TRUE)
grep("Alaska", rawlines, ignore.case = TRUE)
```

Find all 50 state names, using the built-in state data. 

```{r}
state.name
```


```{r}
state_name <- state.name
indices <- sapply(state_name, function(x) grep(x, rawlines, ignore.case = TRUE, fixed = FALSE))
str(indices)
```

Some have two results because "kansas" as a string is also in Arkansas. Virginia is also in West Virginia. washingtom is in the footnotes

```{r}
indices$Kansas     <- indices$Kansas[2]
indices$Virginia   <- indices$Virginia[1]
indices$Washington <- indices$Washington[1]
str(indices)
```

```{r}
first_row <- unname(unlist(indices))
str(first_row)

delta <- diff(first_row)
delta <- diff(first_row)[1] - 3
delta
```

Construct data frame with indices

```{r}
df_indices <- data_frame(state = state_name) %>% 
	mutate(first_row = first_row) %>% 
	mutate(last_row = first_row + delta)



glimpse(df_indices)
```

Extract one state and write to file

```{r}
j = 6
this_state <- df_indices$state[j]
state_data <- rawlines[df_indices$first_row[j]:df_indices$last_row[j]]
state_data

this_path <- paste0("data/", this_state, ".csv")
write_lines(state_data, this_path)
```

Read the file , keep only the columns we need

```{r}
df <- read_csv(this_path, skip = 2, col_names = FALSE) %>% 
	mutate(state = this_state) 

kable(df)
```

At this point I have accomplished just about the same as I did earlier using jailbreakr. 




##  tidying the data for one state

Whether I used jailbreakr or readr, I have the data for one state in a form that can be manipulated to find the data i need. 

Extract the first row (total number of households) and last row (median income). Each column represents a different household size: total households, 1-person, 2-person, 3-person, etc. 

```{r}
m = dim(df)[1]
N_households   <- df[1, 2:8]
median_dollars <- df[m, 2:8]


N_households
median_dollars
```

Summarize the number of households total and the median income. 

```{r}
N_households <- unname(unlist(N_households[1, 1]))
N_households
median_dollars <- unname(unlist(median_dollars[1, 1]))
median_dollars
```

Omit the first and last row, add the median column 

```{r}
df <- df[2:(m-1), ] %>% 
	mutate(median_income =  median_dollars) %>% 
	select(state, X2, median_income, X1) %>% 
	rename(bracket = X1, households = X2)

m = dim(df)[1]
knitr::kable(df)
```




Separate the brackets

```{r}
df <- df %>%
	separate(bracket, into = c("lower", "upper"), sep = "to")

knitr::kable(df)
```

Move the first row "less than" 

```{r}
df <- df %>% 
	mutate(lower = str_replace(lower, "Less than", "")) %>% 
	mutate(lower = str_replace(lower, "or more", ""))

df$upper[1] <- df$lower[1]
df$lower[1] <- 0
df$upper[m] <- NA

knitr::kable(df)
```

Now remove all the dollar signs

```{r}
df <- df %>% 
	mutate(lower = str_replace(lower, "\\$", "")) %>% 
	mutate(lower = str_replace(lower, ",", "")) %>% 
	mutate(lower = as.integer(lower))

df <- df %>% 
	mutate(upper = str_replace(upper, "\\$", "")) %>% 
	mutate(upper = str_replace(upper, ",", "")) %>% 
	mutate(upper = as.integer(upper))

glimpse(df)

knitr::kable(df)
```


The data are now tidy and can be used to compute the income inequality values I'm looking for. 

This is a point at which i would write the state data to file and repeat (using a for loop for example) for all 50 states. 

## analysis 

My goal is to determine income inequality of the state, defined as the ratio of the total income of all households earning the state median income or less to the total income of all households in the state. 



The max bracket can't be averaged because there is no upper limit, so I'll just assign lower to upper for that row. 

```{r}
df$upper[m] <- df$lower[m]
```


use the mean of the bracket to estimate the income for that bracket's households. 

```{r}
df <- df %>%
	mutate(bracket_mean = round((lower + upper) / 2, 0))

knitr::kable(df)
```

Finally, the product of mean income and the number of households gives a total income for each bracket. Use these results to find the total income. 

```{r}
df <- df %>% 
	mutate(bracket_income = households * bracket_mean)

knitr::kable(df)
```

Group and summarize

```{r}
all_income <- df %>% 
	group_by(state) %>% 
	summarize(all_income_M = sum(bracket_income) / 1e+6)

all_income 
```

Filter for all households with income less than or equal to the median. Use these results to find the total income of low-income households. Income is now in millions of dollars. 

```{r}
low_income <- df %>% 
	filter(lower <= median_income)%>% 
	group_by(state) %>% 
	summarize(low_income_M = sum(bracket_income) / 1e+6)

low_income
```

Join the two df

```{r}
inequality_df <- left_join(low_income, all_income, by = "state") %>% 
	mutate(ineq = round(low_income_M / all_income_M, 3))

kable(inequality_df)
```

Write this to file or add it as a new row to a data frame used for collecting each state or province in a loop. 
