---
output: github_document
bibliography: "../resources/bib-etc/references-rr.bib"
csl: "../resources/styles/apa.csl"
link-citations: yes
--- 

# strings

Based on [Chapter 14](http://r4ds.had.co.nz/strings.html) in [@wickhamgrolemund2016].

```{r setup, include = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE)
```


```{r}
library(tidyverse)
library(stringr)
```


I recommend a focus on these sections: 

- 14.2.1 `str_length()`   
- 14.2.2 `str_c()`  
- 14.2.3 `str_sub()` 
- 14.3.1 basic matches using regular expressions 
- 14.3.2 anchor the regular expression 
- 14.4.1 `str_detect()` 
- 14.4.5 `str_replace()` 
- 14.4.6 `str_split()` 




## basics 

```{r}
string1 <- "This is a string"
string2 <- 'If I want to include a "quote" inside a string, I use single quotes'

string1

string2
```

Or juse a backslash to inmclude the string 

```{r}
string3 <- "Or I can use a backslash to \"escape\" the quotes inside a string"

string3
```

Note that the output of the string shows the backslashes. To see just the string, use `writeLines()`

```{r}
writeLines(string3)
```

## special characters 

- `\n` newline 
- `\t` tab 
- `\\` backslash 
- `\'` single quote (apostrophe) 
- `\"` quote mark 

Unicode character codes when used in strings in R will print special characters on all platforms. This is just a sample, more codes can be found [here](https://www.rapidtables.com/code/text/unicode-characters.html). 

- `\u03B1`  `r "\u03B1"` 
- `\u03B2`  `r "\u03B2"` 
- `\u20AC`  `r "\u20AC"` 
- `\u00A5`  `r "\u00A5"` 
- `\u00A9`  `r "\u00A9"` 

## combine strings

Combine strings with `str_c()`. 

```{r}
str_c("x", "y")

str_c("x", "y", "z")

```

If you ant them separated, use the `sep` arguments 

```{r}
str_c("x", "y", sep = ", ")
```

Vectorized 

```{r}
str_c("prefix-", c("a", "b", "c"), "-suffix")
```



To collapse a vector of strings into a single string, use `collapse`:

```{r}
a <- c("x", "y", "z")

length(a)

b <- str_c(a, collapse = ", ")

b

length(b)
```

## collapse a string with `|` (or) to search 

Collapse is also useful for creating a string for OR searches. For example, I can create a searchable state data frame, 

```{r}
df <- as.data.frame(state.x77)
df$State <- row.names(state.x77)
df <- as_tibble(df)
df
```

Then use collapse to create a search string with the OR operator `|` 

```{r}
search_terms <- c("California", "Oregon", "Washington")
search_string <- str_c(search_terms, collapse = "|")
search_string 
```

Then filter by detecting any of the terms in any column. `filter_all()` and `any_vars()` are dplyr functions. 

```{r}
filter_all(df, any_vars(str_detect(., search_string)))
```

Note also the use of `.` as a placeholder for `df` 












## references
