---
output: github_document
bibliography: "../resources/bib-etc/references-rr.bib"
csl: "../resources/styles/apa.csl"
--- 


# multiway graphs

```{r setup, echo = FALSE}
library(knitr)
opts_knit$set(root.dir = "../")
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, collapse = TRUE, fig.keep = 'high')
opts_chunk$set(fig.height=6)
```


```{r echo = FALSE}
library(tidyverse)
library(readxl)
library(stringr)
```

Multiway data have a specific structure

- a categorical variable with *m* levels
- a second categorical variable with *n* levels 
- a quantitative variable with  *m x n* values, one value for each combination of the two categories


## data

Create `multiway.Rmd` in your `practice_work/exercises/` directory. Write the code chunks in the tutorial with as much of the prose as you like  to explain the work. 

```{r}
library(tidyverse)
library(forcats)
```

An untidy data file `midfield-student-data.csv` adapted from [@layton2009] is saved to the course data directory on GitHub. You can download the file and save it in your `practice_work/data/` directory. Once the file is saved on your machine, read the file with your Rmd script. 

```{r}
df1 <- read_csv('data/midfield-student-data.csv')

glimpse(df1)
```

Multiway data are often displayed in a wide table. The *m* levels of the first category are in the rows of first column; the *n* levels of the second category are in the other column headings; and the *m x n* values of the quantitative variable fill the table cells---no cell is empty. 

```{r}
kable(df1)
```

Categorical variable 1---a student's major after 4 years---has 8 levels: 

- Arts and Humanities
- Business
- Computer Science
- Engineering
- Other Non-STM disciplines (STM: Science, Technology, and Math)
- Other STM disciplines
- Social Sciences
- TOLEDO: Destination unknown (left the dataset)

Categorical variable 2---a student's race/ethinicty and sex---has 10 levels: 

- Asian American Female and Male
- Black (African American) Female and Male
- Hispanic (Latin) Female and Male
- Native American Female and Male
- White (non-Hispanic) Female and Male

Quantitative variable has 8 x 10 = 80 values 

- Number of students enrolled in a major for each race/ethnicity/sex group 



## tidying the data 

Use *gather()* to tidy the data (wide form to long form).

```{r}
df2 <- df1 %>%
	gather(demogr, N, AsianFem:WhiteMle)

glimpse(df2)
```

Save the tidy data as a new CSV file in your `data/` directory. 

```{r}
write_csv(df2, 'data/midfield-student-data-tidy.csv')
```




## graphing multiway data

Read the tidy CSV data file from your data directory.

```{r}
df3 <- read_csv('data/midfield-student-data-tidy.csv')

glimpse(df3)
```

Plot it with rows (y) the demographic category and facets the destination major category. The x-variable is the number of students N. 


```{r}
f1 <- ggplot(df3, aes(x = N, y = demogr)) +
	geom_point() +
	facet_wrap(vars(DestMajor), ncol = 3)

print(f1)
```

The multiway categorical variables must be recoded as factors that are ordered by N. 

```{r}
df3 <- df3 %>% 
  mutate(demogr = fct_reorder(demogr, N, sum)) %>%
  mutate(DestMajor = fct_reorder(DestMajor, N, sum))
```

```{r}
f3 <- ggplot(df3, aes(x = N, y = demogr)) +
  geom_point() +
  facet_wrap(vars(DestMajor), ncol = 3, as.table = FALSE)

print(f3)
```





## applying a log scale

With data spanning several orders of magnitude, a logarithmic scale let us see small and large differences.

```{r}
f3 <- f3 + scale_x_log10()
print(f3)
```

Your console may show a warning message about a transformation error. 

```
Warning message:
Transformation introduced infinite values in continuous x-axis 
```

The warning was issued because I tried to graph log(0) which is undefined. We can filter the data to keep only those rows with $N > 0$.

```{r}
df4 <- df3 %>%
  filter(N > 0)

# how many observations did we lose?
glimpse(df4)
```






## using %+% to replace a current data frame

We want to use the data frame `df4` that has been filtered to remove entries with $N = 0$.

We replace a data frame in a ggplot graph we've already defined using the `%+%` operator---similar to using `+` to add a new layer.

```{r}
f4 <- f3 %+% df4 
print(f4)
```




## recoding row and panel labels

In preparing the graph for sharng with others, we will typically replace the categorical levels with easier to read labels. Following the suggestion is [section 15.5](http://r4ds.had.co.nz/factors.html#modifying-factor-levels) of the text [@wickhamgrolemund2016]:


```{r}
df5 <- df4 %>% 
  mutate(demogr = fct_recode(demogr,
    "Asian Male"           = "AsianMle",
    "Asian Female"         = "AsianFem",
    "Black Male"           = "BlackMle",
    "Black Female"         = "BlackFem",
    "Hispanic Male"        = "HispMle",
    "Hispanic Female"      = "HispFem",
    "Native American Male" = "NatAmMle",
    "Native American Female" = "NatAmFem", 
    "White Male"           = "WhiteMle",
    "White Female"         = "WhiteFem"
  )) %>% 
  mutate(DestMajor = fct_recode(DestMajor,
    "Arts &Humanities"  = "AH",
    "Business"          = "Bus",
    "Computer Science"  = "CS",
    "Engineering"       = "Engr",
    "Other Non-STM"     = "otherNonSTM",
    "Other STM"         = "otherSTM",
    "Left the database" = "TOLEDO",
    "Social Sciences"   = "SocSci", 
  ))

# replace df and add labels 
f5 <- f4 %+%
	df5 +
  labs(x = "Number of students (log 10 scale)", 
      y = "", 
      title = "The year-4 destinations of students starting in engineering")
	
print(f5)
```





## the dual multiway

In a multiway graph, numbers in a panel are more easily compared than numbers in different panels. 

Every multiway has dual form in which we swap the categories---rows become panels and panels become rows. All we have to do is change the variables assigned to `y` and `facet`. 


```{r fig.height=7.5}
f6 <- ggplot(df5, aes(x = N, y = DestMajor)) +
	geom_point() +
	facet_wrap(vars(demogr), ncol = 3, as.table = FALSE) +
	scale_x_log10() + 
	labs(x = "Number of students (log 10 scale)", 
      y = "", 
      title = "The year-4 destinations of students starting in engineering")

print(f6)
```










## references

<div id="refs"></div>



<!-- other stuff can go after -->


