---
output: github_document
---

```{r setup, include=FALSE}
library(knitr)
opts_chunk$set(echo = TRUE)
opts_chunk$set(collapse = TRUE)
```

# Ch 1 examples and exercises

## examples

Load the tidyverse packages

```{r}
library(tidyverse)
```

Examine the `mpg` data set

```{r}
mpg
```

First graph

```{r}
ggplot(data = mpg) +
	geom_point(mapping = aes(x = displ, y = hwy))
```

## exercises, p.6 


1. Produces an empty graph because no geometric object has been specified. 

```{r}
ggplot(data = mpg)
```

2. Rows and columns in `mpg` from the output below 234 rows, 11 columns.  

```{r}
mpg
```

3. What does the drv variable describe? Read the help for ?mpg to find out.

- f = front-wheel drive
- r = rear wheel drive
- 4 = 4wd

4. Make a scatterplot of hwy vs cyl.

```{r}
ggplot(data = mpg) +
	geom_point(mapping = aes(x = cyl, y = hwy))
```


5. What happens if you make a scatterplot of class vs drv? Why is the plot not useful?

```{r}
ggplot(data = mpg) +
	geom_point(mapping = aes(x = drv, y = class))
```

The plot is not particularly useful because class is generally independent of drive type, so no correlation between the two variables is seen. 





