---
output: github_document
bibliography: "../../resources/bib-etc/references-rr.bib"
csl: "../../resources/styles/apa.csl"
--- 


# Case study in tidying data

From section  [12.6](http://r4ds.had.co.nz/tidy-data.html#case-study) in R for Data Science. 


The `who` dataset is in the tidyr package, part of the tidyverse

```{r}
library(tidyverse)
glimpse(who)
```

- The first three apprear to be redundant encodings of the country name. 
- Year appears to be a variable 
- The remnaining column headings appear to encode another variable 

Type `? who` in the Console to see what the authors state are the variables. 

```
? who
```

The data under the encoded columns are "Counts of new TB cases recorded by group". 

```{r}
who1 <- who %>% 
  gather(new_sp_m014:newrel_f65, key = "key", value = "cases", na.rm = TRUE)

who1
```

Use count to assess the new key column

```{r}
who1 %>% 
  count(key)
```


From the help page: 


> The data uses the original codes given by the World Health Organization. 
>
> The column names for columns five through 60 are made by combining new_ to a code for method of diagnosis new = new cases, old = old cases    
> 
> type of TB 
rel = relapse    
sn = negative pulmonary smear    
sp = positive pulmonary smear    
ep = extrapulmonary    
>
> Sex   
f = female    
m = male    
>
> Age group   
014 = 0-14 yrs of age    
1524 = 15-24 years of age    
2534 = 25 to 34 years of age    
3544 = 35 to 44 years of age    
4554 = 45 to 54 years of age    
5564 = 55 to 64 years of age    
65 = 65 years of age or older    


Let's look for unique values of "key". We find an error in the data encoding

```{r}
who1 %>% 
  count(key) %>% 
  select(key) %>% 
  as.data.frame() %>% 
  unique() %>% 
  print()
```



```{r}
who2 <- who1 %>% 
  mutate(key = stringr::str_replace(key, "newrel", "new_rel"))

who2
```

We can separate the values in each code with two passes of separate(). The first pass will split the codes at each underscore.

```{r}
who3 <- who2 %>% 
  separate(key, c("new", "type", "sexage"), sep = "_")

who3
```

Then we might as well drop the new column because it’s constant in this dataset. 


```{r}
who3 %>% 
  count(new)
```

While we’re dropping columns, let’s also drop iso2 and iso3 since they’re redundant.


```{r}
who4 <- who3 %>% 
  select(-new, -iso2, -iso3)

glimpse(who4)
```

Next we’ll separate sexage into sex and age by splitting after the first character:


```{r}
who5 <- who4 %>% 
  separate(sexage, c("sex", "age"), sep = 1)

who5
```

Lastly, we'rte in tidy form, let's recode the entries.  

```{r}
who6 <- who5 %>% 
	mutate(
    sex = recode(sex
      , "m" = "male"
      , "f" = "female"
    )
	) %>% 
	mutate(
    type = recode(type
      , "sp" = "pos pulm smear"
      , "sn" = "neg pulm smear"
      , "ep" = "extrapulmonary"
    )
	)	


who6
```


Summary

```{r}
summary(who6)
```

```{r}
who7 <- filter(who6, cases > 0)
ggplot(who7, aes(x = cases)) +
	geom_density() +
	scale_x_log10(breaks = 10^c(0:5)) +
	geom_vline(data = who7, mapping = aes(xintercept = median(cases)))
```

